---
title: "Alcohol Dose and Mental Arithmetic Performance"
subtitle: "A Randomized Block Experiment on The Islands"
author: "STATS 101B, Group 29 --- Jack Lutz, Kathy Nguyen, et al."
date: "June 7, 2026"
output:
  pdf_document:
    number_sections: true
    fig_caption: true
geometry: margin=0.9in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 3.2, fig.align = "center")
library(ggplot2)
library(pwr)
library(knitr)

navy <- "#1E2761"; terra <- "#B85042"; ink <- "#1A1A1A"; gray <- "#6B7280"

d <- read.csv("../data/between_subjects.csv", stringsAsFactors = FALSE)
d$baseline   <- ifelse(d$difficulty == "Easy", d$baseline_easy, d$baseline_hard)
stopifnot(all(d$baseline - d$response == d$diff))   # data integrity check
d$difficulty <- factor(d$difficulty, levels = c("Easy", "Hard"))
d$dose       <- factor(d$shots, levels = c(1, 2, 3))
d$shots_n    <- d$shots

# --- models ---------------------------------------------------------------
fit      <- aov(diff ~ difficulty * dose, data = d)   # full factorial / GRBD
fit_add  <- aov(diff ~ difficulty + dose, data = d)   # additive RBD
fit_lin  <- lm(diff ~ difficulty * shots_n, data = d) # dose as ordinal trend
an       <- summary(fit)[[1]]

getp <- function(term) signif(an[trimws(rownames(an)) == term, "Pr(>F)"], 2)
getF <- function(term) round(an[trimws(rownames(an)) == term, "F value"], 2)

# within-difficulty dose slopes
sl_easy <- coef(lm(diff ~ shots_n, data = subset(d, difficulty == "Easy")))[2]
sl_hard <- summary(lm(diff ~ shots_n, data = subset(d, difficulty == "Hard")))$coef

# effect size (Cohen's f) for the dose main effect
ss   <- an[, "Sum Sq"]; names(ss) <- trimws(rownames(an))
f_dose <- sqrt((ss["dose"] / (ss["dose"] + ss["Residuals"])) /
               (1 - ss["dose"] / (ss["dose"] + ss["Residuals"])))

n_large <- pwr.anova.test(k = 3, f = 0.40, sig.level = 0.05, power = 0.80)$n
n_obs   <- pwr.anova.test(k = 3, f = as.numeric(f_dose), sig.level = 0.05,
                          power = 0.80)$n
pow_20  <- pwr.anova.test(k = 3, n = 20, f = 0.40, sig.level = 0.05)$power
```

# Introduction

Alcohol consumption is a common social behavior, especially in the college-age
demographic, and it has the potential to influence cognitive functions such as
memory, attention, and problem solving. Because mental arithmetic underlies many
everyday and academic tasks, it is worth asking how alcohol affects it and
whether the **amount** consumed matters. Our investigative questions are:
(1) does drinking alcohol change performance on a timed mental-arithmetic task,
and (2) does the size of that change depend on the dose and on the difficulty of
the task?

Going in, we expected alcohol to reduce performance, and we expected the
reduction to grow with dose. We also suspected the effect would be larger on
harder problems, which demand more working memory and controlled attention,
than on easy, well-practiced problems. We tested these expectations with a
designed experiment on the virtual population of **The Islands**
(Providence Island), where each inhabitant can be sampled, given a measured dose
of alcohol, and timed on a standardized arithmetic task.

# Design of the Experiment

## From the original design to a corrected one

Our first design gave **every subject all six** alcohol-by-difficulty
conditions in a single session, treating each subject as a block. This is not a
valid randomized block design for alcohol: doses were administered in sequence
with no washout, so cumulative dose, fatigue, and practice on the task are
confounded with the alcohol level, and the handful of subjects could not support
generalization. We therefore redesigned the study as a **between-subjects**
experiment: each islander is measured once at a sober **baseline**, then
randomly assigned to exactly **one** treatment and re-tested.

## Response: change from baseline (a proxy for math skill)

Islanders differ enormously in baseline arithmetic ability, and that ability is
not the thing we want to measure. Rather than analyze raw post-treatment scores,
we use each subject's own sober baseline as a proxy for their underlying math
skill and analyze the **change**,
$$D = (\text{baseline score}) - (\text{score after treatment}),$$
the number of correct answers *lost* relative to that subject's own sober
performance at the same difficulty. Differencing against the personal baseline
removes between-subject ability so the analysis isolates the alcohol-induced
change.

## Treatments, blocks, and randomization

This is a **randomized block design** with a $2 \times 3$ treatment layout,
analyzed with replication (a generalized RBD):

- **Treatment factor --- alcohol dose:** 1, 2, or 3 shots of vodka (30 mL each).
  The sober baseline supplies the no-alcohol reference, so no separate 0-shot
  group is needed.
- **Blocking factor --- task difficulty:** Easy or Hard. Difficulty is the
  largest known source of variation in performance change, so we block on it to
  estimate the dose effect more precisely; replication within blocks also lets
  us test whether the dose effect differs by difficulty.
- **Experimental units:** islanders aged 21+, randomly sampled from households on
  Providence Island and randomly assigned to a dose within each difficulty
  block. Each unit receives one treatment, eliminating the carryover that
  invalidated the original design.

## Sample size and power

We fixed the sample size in advance with a power analysis. Treating dose as a
three-level factor, detecting a **large** effect (Cohen's $f = 0.40$;
Cohen, 1988) at $\alpha = 0.05$ with power $0.80$ requires
`r ceiling(n_large)` islanders per dose level:

```{r power, echo=TRUE}
pwr.anova.test(k = 3, f = 0.40, sig.level = 0.05, power = 0.80)$n
```

We collected **`r nrow(d)` islanders**, balanced as **10 per cell**
(20 per dose, 30 per difficulty block), giving power
`r round(pow_20, 2)` for a large dose effect and effectively full power for the
large difficulty effect. The design is completely balanced, so all factor
effects are orthogonal.

# Results and Interpretation

Table 1 and Figure 1 summarize the mean drop $D$ in each cell. The pattern is
striking: on Easy tasks the drop hovers near one point regardless of dose,
whereas on Hard tasks it climbs steadily from `r round(mean(d$diff[d$difficulty=="Hard" & d$shots_n==1]),1)`
at one shot to `r round(mean(d$diff[d$difficulty=="Hard" & d$shots_n==3]),1)` at three.

```{r celltable}
cm <- tapply(d$diff, list(d$difficulty, d$dose), mean)
tab <- as.data.frame(round(cm, 2))
colnames(tab) <- c("1 shot", "2 shots", "3 shots")
kable(tab, caption = "Mean drop in score $D$ (baseline $-$ post) by difficulty block and dose. Each cell is 10 islanders.")
```

```{r interaction, fig.height=2.9, fig.cap="Mean drop in arithmetic score (with $\\pm 1$ SE) by alcohol dose, separately for Easy and Hard tasks. Alcohol degrades hard-task performance in a dose-dependent way; easy-task performance is unaffected."}
agg <- aggregate(diff ~ difficulty + dose, d,
                 function(v) c(m = mean(v), se = sd(v)/sqrt(length(v))))
agg <- do.call(data.frame, agg); names(agg)[3:4] <- c("m", "se")
ggplot(agg, aes(dose, m, color = difficulty, group = difficulty)) +
  geom_line(linewidth = 1.2) +
  geom_errorbar(aes(ymin = m - se, ymax = m + se), width = .08, linewidth = .8) +
  geom_point(size = 3) +
  scale_color_manual(values = c(Easy = navy, Hard = terra), name = "Task") +
  scale_x_discrete(labels = c("1 shot", "2 shots", "3 shots")) +
  labs(x = "Alcohol dose", y = "Mean drop in score") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        axis.title = element_text(face = "bold", color = ink))
```

## Randomized block ANOVA

```{r anova, echo=TRUE}
fit <- aov(diff ~ difficulty * dose, data = d)   # block * treatment
summary(fit)
```

The blocking factor, **difficulty**, is highly significant
($F_{1,54} = `r getF("difficulty")`$, $p = `r getp("difficulty")`$): hard tasks
lose far more ground under alcohol than easy ones, which confirms that blocking
on difficulty was the right choice. The **dose** main effect averaged over both
blocks is not significant ($F_{2,54} = `r getF("dose")`$,
$p = `r getp("dose")`$), and neither is the block-by-treatment interaction
($p = `r getp("difficulty:dose")`$) as an omnibus test --- both are diluted by
the flat Easy block.

## The dose effect is carried by hard tasks

Because dose is ordered, a linear trend test is the more powerful and direct way
to ask "does the amount matter." Fitting dose as a numeric predictor:

```{r trend, echo=TRUE}
fit_lin <- lm(diff ~ difficulty * shots_n, data = d)
round(coef(summary(fit_lin)), 3)
```

The model explains `r round(summary(fit_lin)$r.squared*100)`% of the variance in
$D$ ($p = `r signif(pf(summary(fit_lin)$fstatistic[1], summary(fit_lin)$fstatistic[2], summary(fit_lin)$fstatistic[3], lower.tail=FALSE), 2)`$).
Within Easy tasks the dose slope is essentially zero
(`r sprintf("%+.2f", sl_easy)` points/shot); within Hard tasks each additional
shot costs **`r sprintf("%.2f", sl_hard[2,1])` correct answers**
(SE `r sprintf("%.2f", sl_hard[2,2])`, two-sided $p = `r round(sl_hard[2,4],3)`$;
one-sided $p = `r round(sl_hard[2,4]/2,3)`$ given our directional prediction).
A Tukey HSD on dose (Appendix) finds no individual pair of doses separable on
its own, consistent with the modest averaged main effect.

## Assumptions

Residuals are consistent with normality (Shapiro-Wilk
$p = `r round(shapiro.test(residuals(fit))$p.value, 2)`$) and with equal
variances across cells (Levene $p = `r round(car::leveneTest(diff ~ difficulty*dose, d)[1,3], 2)`$),
so the ANOVA is well calibrated (diagnostic plots in the Appendix).

# Discussion

**Summary.** Alcohol's effect on mental arithmetic depends on how hard the task
is. Easy, over-learned problems sit near the 40-question ceiling and are
essentially unaffected by one to three shots. Hard problems degrade in a
dose-dependent way, losing roughly `r sprintf("%.1f", sl_hard[2,1])` correct
answers for every additional shot. Task difficulty was the dominant, clearly
significant factor; the dose effect is real but concentrated in the hard block.

**Does this make sense?** Yes. Simple arithmetic is automatic and robust, while
hard arithmetic leans on working memory and controlled attention --- exactly the
faculties that alcohol is known to impair first. The widening Easy-Hard gap with
dose is consistent with the *alcohol-myopia* account (Steele \& Josephs, 1990),
in which alcohol narrows the scope of attention and so disrupts effortful,
attention-demanding cognition more than automatic processing. The flat Easy
curve is a textbook ceiling effect: there is almost no room to fall when sober
performance is already near maximum.

**Limitations.** (1) The Islanders are a simulated population, so external
validity to real drinkers is limited. (2) We powered the study for a large dose
effect; the observed dose effect was smaller ($f \approx `r round(f_dose,2)`$),
so power to detect the dose main effect and the dose-by-difficulty interaction
as significant was low --- reaching 0.80 for an effect this size would need
roughly `r ceiling(n_obs)` islanders per dose (about `r 3*ceiling(n_obs)` total).
(3) The ceiling on Easy tasks compresses its variance and limits sensitivity
there. (4) Change scores inherit measurement error from a single baseline
session. A natural next step is to concentrate sampling on hard tasks, add a
4-or-more-shot level to extend the dose-response curve, and replace the change
score with a baseline-adjusted ANCOVA model.

# References

\small

Cohen, J. (1988). *Statistical Power Analysis for the Behavioral Sciences*
(2nd ed.). Lawrence Erlbaum Associates.

Steele, C. M., \& Josephs, R. A. (1990). Alcohol myopia: Its prized and
dangerous effects. *American Psychologist, 45*(8), 921--933.

\normalsize

\appendix

# Appendix

```{r tukey, echo=TRUE}
TukeyHSD(fit, "dose")$dose
```

```{r diag, echo=FALSE, fig.height=2.7, fig.cap="Diagnostic plots: residual Q-Q and residuals vs. fitted."}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(residuals(fit), main = "Normal Q-Q", pch = 19, col = navy, cex = .6)
qqline(residuals(fit), col = terra, lwd = 2)
plot(fitted(fit), residuals(fit), main = "Residuals vs Fitted",
     xlab = "Fitted", ylab = "Residual", pch = 19, col = navy, cex = .6)
abline(h = 0, lty = 2, col = gray)
```
