---
title: "Alcohol Dose and Mental Arithmetic Performance"
subtitle: "A Randomized Block Experiment on The Islands"
author: "STATS 101B, Group 29 --- Kevin Brady, Jonathan Deng, Sung Heon Kang, Jack Lutz, Kathy Nguyen, Siri Suvvari"
date: "June 7, 2026"
output:
  pdf_document:
    number_sections: true
    fig_caption: true
  word_document:
    fig_caption: true
geometry: margin=0.9in
fontsize: 11pt
params:
  show_code: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 3.2, fig.align = "center")
library(ggplot2); library(pwr); library(knitr); library(car)

navy <- "#1E2761"; terra <- "#B85042"; ink <- "#1A1A1A"; gray <- "#6B7280"

d <- read.csv("../data/between_subjects.csv", stringsAsFactors = FALSE)
d$baseline   <- ifelse(d$difficulty == "Easy", d$baseline_easy, d$baseline_hard)
stopifnot(all(d$baseline - d$response == d$diff))   # data integrity check
d$difficulty <- factor(d$difficulty, levels = c("Easy", "Hard"))
d$dose       <- factor(d$shots, levels = c(1, 2, 3))
d$shots_n    <- d$shots

# --- change-score models (primary) ----------------------------------------
fit     <- aov(diff ~ difficulty * dose, data = d)    # block * treatment
fit_lin <- lm(diff ~ difficulty * shots_n, data = d)  # dose as ordinal trend
an      <- summary(fit)[[1]]
getp <- function(t) signif(an[trimws(rownames(an)) == t, "Pr(>F)"], 2)
getF <- function(t) round(an[trimws(rownames(an)) == t, "F value"], 2)
sl_easy <- coef(lm(diff ~ shots_n, data = subset(d, difficulty == "Easy")))[2]
sl_hard <- summary(lm(diff ~ shots_n, data = subset(d, difficulty == "Hard")))$coef

# --- ANCOVA models (robustness) --------------------------------------------
anc     <- lm(response ~ baseline + difficulty * dose, data = d)
anc_an  <- Anova(anc, type = 2)
anc_lin <- lm(response ~ baseline + difficulty * shots_n, data = d)
b_base  <- coef(anc)["baseline"]
sd_diff <- sigma(fit); sd_anc <- sigma(anc)
anc_dose_p <- anc_an["dose", "Pr(>F)"]
anc_int_p  <- summary(anc_lin)$coef["difficultyHard:shots_n", "Pr(>|t|)"]
hslope_anc <- summary(lm(response ~ baseline + shots_n,
                         data = subset(d, difficulty == "Hard")))$coef["shots_n", ]
slopes_p   <- Anova(lm(response ~ baseline * dose + baseline * difficulty +
                       difficulty * dose, data = d), type = 2)["baseline:dose", "Pr(>F)"]

# --- effect size + power ---------------------------------------------------
ss <- an[, "Sum Sq"]; names(ss) <- trimws(rownames(an))
f_dose <- sqrt((ss["dose"]/(ss["dose"]+ss["Residuals"])) /
               (1 - ss["dose"]/(ss["dose"]+ss["Residuals"])))
n_large <- pwr.anova.test(k = 3, f = 0.40, sig.level = 0.05, power = 0.80)$n
n_obs   <- pwr.anova.test(k = 3, f = as.numeric(f_dose), sig.level = 0.05, power = 0.80)$n
pow_20  <- pwr.anova.test(k = 3, n = 20, f = 0.40, sig.level = 0.05)$power

# --- helpers to print clean tables in both PDF and Word --------------------
fmt_anova <- function(tab) {
  a <- as.data.frame(tab)
  pcol <- grep("Pr", names(a))
  num  <- setdiff(seq_along(a), pcol)
  a[num] <- lapply(a[num], function(x) ifelse(is.na(x), NA, round(x, 2)))
  a[[pcol]] <- ifelse(is.na(a[[pcol]]), "", formatC(a[[pcol]], format = "f", digits = 4))
  a
}
```

# Introduction

Alcohol consumption is a common social behavior, especially in the college-age
demographic, and it has the potential to influence cognitive functions such as
memory, attention, and problem solving. Because mental arithmetic underlies many
everyday and academic tasks, it is worth asking how alcohol affects it and
whether the **amount** consumed matters. Our investigative questions are:
(1) does drinking alcohol affect performance on a timed mental-arithmetic task,
and (2) does the size of that change depend on the dose and on the difficulty of
the task?

Going in, we expected alcohol to reduce performance, and we expected the
reduction to grow with dose. We also suspected the effect would be larger on
harder problems, which demand more working memory and controlled attention,
than on easy, well-practiced problems. We tested these expectations with a
designed experiment on the virtual population of **The Islands**
(Providence Island), where each inhabitant can be sampled, given a measured dose
of alcohol, and timed on a standardized arithmetic task.

# Design of the Experiment

## From the original design to a corrected one

Our first design gave **every subject all six** alcohol-by-difficulty
conditions in a single session, treating each subject as a block. This is not a
valid randomized block design for alcohol: doses were administered in sequence
with no washout, so cumulative dose, fatigue, and practice on the task are
confounded with the alcohol level, and the handful of subjects could not support
generalization. We therefore redesigned the study as a **between-subjects**
experiment: each islander is measured once at a sober **baseline**, then
randomly assigned to exactly **one** treatment and re-tested. Random assignment
is what licenses the causal comparison: any systematic difference between dose
groups is attributable to dose, not to who landed in which group.

## Response: change from baseline (a proxy for math skill)

Islanders differ enormously in baseline arithmetic ability, and that ability is
not the thing we want to measure. Rather than analyze raw post-treatment scores,
we use each subject's own sober baseline as a proxy for their underlying math
skill and analyze the **change**, $D = \text{baseline} - \text{post}$: the
number of correct answers *lost* relative to that subject's own sober
performance at the same difficulty. Differencing against the personal baseline
removes between-subject ability so the analysis isolates the alcohol-induced
change.

## Treatments, blocks, and randomization

This is a **randomized block design** with a $2 \times 3$ treatment layout,
analyzed with replication (a generalized RBD):

- **Treatment factor --- alcohol dose:** 1, 2, or 3 shots of vodka (30 mL each).
  The sober baseline supplies the no-alcohol reference, so no separate 0-shot
  group is needed.
- **Blocking factor --- task difficulty:** Easy or Hard. Difficulty is the
  largest known source of variation in performance change, so we block on it to
  estimate the dose effect more precisely; replication within blocks also lets
  us test whether the dose effect differs by difficulty.
- **Experimental units:** islanders aged 21+, randomly sampled from households on
  Providence Island and randomly assigned to a dose within each difficulty
  block. Each unit receives one treatment, eliminating the carryover that
  invalidated the original design.

## Sample size and power

We fixed the sample size in advance with a power analysis. Treating dose as a
three-level factor, detecting a **large** effect (Cohen's $f = 0.40$;
Cohen, 1988) at $\alpha = 0.05$ with power $0.80$ requires
`r ceiling(n_large)` islanders per dose level:

```{r power, echo=params$show_code}
pwr.anova.test(k = 3, f = 0.40, sig.level = 0.05, power = 0.80)$n
```

We collected **`r nrow(d)` islanders**, balanced as **10 per cell**
(20 per dose, 30 per difficulty block), giving power
`r round(pow_20, 2)` for a large dose effect and effectively full power for the
large difficulty effect. The design is completely balanced, so all factor
effects are orthogonal. As we report below, the realized dose effect was smaller
than the large one we planned for, a point we return to in the Discussion.

# Results and Interpretation

Table 1 and Figure 1 summarize the mean drop $D$ in each cell. The pattern is
striking: on Easy tasks the drop hovers near one point regardless of dose,
whereas on Hard tasks it climbs steadily from `r round(mean(d$diff[d$difficulty=="Hard" & d$shots_n==1]),1)`
at one shot to `r round(mean(d$diff[d$difficulty=="Hard" & d$shots_n==3]),1)` at three.

```{r celltable}
cm <- tapply(d$diff, list(d$difficulty, d$dose), mean)
tab <- as.data.frame(round(cm, 2)); colnames(tab) <- c("1 shot", "2 shots", "3 shots")
kable(tab, caption = "Mean drop in score (sober baseline minus post-treatment) by difficulty block and dose. Each cell is 10 islanders.")
```

```{r interaction, fig.height=2.9, fig.cap="Mean drop in arithmetic score (with one-standard-error bars) by alcohol dose, for Easy and Hard tasks. Alcohol degrades hard-task performance in a dose-dependent way; easy-task performance is essentially unaffected."}
agg <- aggregate(diff ~ difficulty + dose, d,
                 function(v) c(m = mean(v), se = sd(v)/sqrt(length(v))))
agg <- do.call(data.frame, agg); names(agg)[3:4] <- c("m", "se")
ggplot(agg, aes(dose, m, color = difficulty, group = difficulty)) +
  geom_line(linewidth = 1.2) +
  geom_errorbar(aes(ymin = m - se, ymax = m + se), width = .08, linewidth = .8) +
  geom_point(size = 3) +
  scale_color_manual(values = c(Easy = navy, Hard = terra), name = "Task") +
  scale_x_discrete(labels = c("1 shot", "2 shots", "3 shots")) +
  labs(x = "Alcohol dose", y = "Mean drop in score") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        axis.title = element_text(face = "bold", color = ink))
```

## Primary analysis: randomized block ANOVA

```{r anovacode, echo=params$show_code}
fit <- aov(diff ~ difficulty * dose, data = d)   # block * treatment, change score
```
```{r anovatab}
kable(fmt_anova(summary(fit)[[1]]),
      caption = "Change-score ANOVA. Difficulty is the blocking factor, dose the treatment.")
```

The blocking factor, **difficulty**, is highly significant
($F_{1,54} = `r getF("difficulty")`$, $p = `r getp("difficulty")`$): hard tasks
lose far more ground under alcohol than easy ones, which confirms that blocking
on difficulty was the right choice. The **dose** main effect averaged over both
blocks is not significant ($F_{2,54} = `r getF("dose")`$,
$p = `r getp("dose")`$), and neither is the block-by-treatment interaction
($p = `r getp("difficulty:dose")`$) as an omnibus test: both are diluted by the
flat Easy block.

## The dose effect is carried by hard tasks

Because dose is ordered, a planned linear trend test is the more powerful and
direct way to ask "does the amount matter."

```{r trend, echo=params$show_code}
fit_lin <- lm(diff ~ difficulty * shots_n, data = d)  # dose as a numeric trend
```

The trend model explains `r round(summary(fit_lin)$r.squared*100)`% of the
variance in $D$ ($p = `r signif(pf(summary(fit_lin)$fstatistic[1], summary(fit_lin)$fstatistic[2], summary(fit_lin)$fstatistic[3], lower.tail=FALSE), 2)`$).
Within Easy tasks the dose slope is essentially zero
(`r sprintf("%+.2f", sl_easy)` points/shot); within Hard tasks each additional
shot costs **`r sprintf("%.2f", sl_hard[2,1])` correct answers**
(SE `r sprintf("%.2f", sl_hard[2,2])`, two-sided $p = `r round(sl_hard[2,4],3)`$;
one-sided $p = `r round(sl_hard[2,4]/2,3)`$ given our directional prediction).

## Robustness: analysis of covariance

The change score implicitly assumes a one-point gain in sober baseline maps to a
one-point gain in treated score. We relax that by regressing the raw
post-treatment score on the baseline, which also corrects for regression to the
mean:

```{r ancovacode, echo=params$show_code}
anc <- lm(response ~ baseline + difficulty * dose, data = d)  # ANCOVA
```
```{r ancovatab}
kable(fmt_anova(Anova(anc, type = 2)),
      caption = "ANCOVA (Type II): post-treatment score adjusted for the sober baseline.")
```

The fitted baseline coefficient is `r round(b_base,2)`, close to the 1.0 the
change score imposes, so the two models agree and the residual scatter barely
moves (`r round(sd_anc,2)` vs. `r round(sd_diff,2)`). Baseline ability and
difficulty dominate, and the omnibus dose effect stays non-significant
($p = `r round(anc_dose_p,2)`$). The one shift is that the planned hard-task dose
contrast sharpens to `r round(hslope_anc[1],2)` correct per shot
(two-sided $p = `r round(hslope_anc[4],3)`$, one-sided $`r round(hslope_anc[4]/2,3)`$),
and the difficulty-by-dose interaction moves to $p = `r round(anc_int_p,3)`$.
Because the baseline slope varies modestly across doses (equal-slopes test
$p = `r round(slopes_p,2)`$), we report the ANCOVA **alongside**, not in place
of, the change-score model. Both lead to the same conclusion.

## Assumptions

Residuals are consistent with normality (Shapiro-Wilk
$p = `r round(shapiro.test(residuals(fit))$p.value, 2)`$) and with equal
variances across cells (Levene $p = `r round(car::leveneTest(diff ~ difficulty*dose, d)[1,3], 2)`$),
so the ANOVA is well calibrated (diagnostic plots in the Appendix).

# Discussion

**Summary.** Alcohol's effect on mental arithmetic depends on how hard the task
is. Easy, over-learned problems sit near the 40-question ceiling and are
essentially unaffected by one to three shots. Hard problems degrade in a
dose-dependent way, losing roughly `r sprintf("%.1f", sl_hard[2,1])` correct
answers for every additional shot (significant under ANCOVA, marginal under the
change score). Task difficulty was the dominant, clearly significant factor; the
dose effect is real but concentrated in the hard block.

**Does this make sense?** Yes. Simple arithmetic is automatic and robust, while
hard arithmetic leans on working memory and controlled attention --- exactly the
faculties that alcohol is known to impair first. The widening Easy-Hard gap with
dose is consistent with the *alcohol-myopia* account (Steele \& Josephs, 1990),
in which alcohol narrows the scope of attention and so disrupts effortful,
attention-demanding cognition more than automatic processing. The flat Easy
curve is a textbook ceiling effect: there is almost no room to fall when sober
performance is already near maximum.

**Limitations.**

1. *Simulated population.* The Islanders are a virtual population, so the effects
   reflect the simulation's model and external validity to real drinkers is
   limited; the sample is also drawn from a single island.
2. *Practice (warm-up) confound.* Every subject took the sober baseline before
   the treated test, so practice warms them up and inflates the post-treatment
   score, which shrinks $D$. This biases the estimated *absolute* impairment
   toward zero. Because dose was randomized, the practice gain is balanced across
   dose groups and largely cancels in the dose-to-dose comparison, but a sober
   test--retest control arm would be needed to net it out cleanly.
3. *Dose is not exposure.* Shots were not titrated to body water (weight, sex) or
   tolerance, so a fixed dose produces different blood-alcohol levels and
   different impairment across people. This heterogeneity inflates within-cell
   variance and is part of why the study is underpowered.
4. *Underpowered for the realized effect.* We powered for a large effect
   ($f = 0.40$); the observed dose effect was smaller ($f \approx `r round(f_dose,2)`$),
   so detecting it at $0.80$ power would need roughly `r ceiling(n_obs)`
   islanders per dose (about `r 3*ceiling(n_obs)` total).
5. *Ceiling on easy tasks.* Sober easy scores sit at 38--40 of 40, leaving almost
   no room to fall, so a genuine easy-task effect could be masked.

Natural next steps follow directly: add a sober test--retest control arm to
remove practice, record weight and sex to model blood-alcohol concentration,
concentrate sampling on hard tasks, and extend the dose range beyond three shots.

# References

Cohen, J. (1988). *Statistical Power Analysis for the Behavioral Sciences*
(2nd ed.). Lawrence Erlbaum Associates.

Steele, C. M., \& Josephs, R. A. (1990). Alcohol myopia: Its prized and
dangerous effects. *American Psychologist, 45*(8), 921--933.

\appendix

# Appendix

```{r tukey, echo=params$show_code}
TukeyHSD(fit, "dose")$dose   # pairwise dose comparisons (change score)
```

```{r diag, fig.height=2.7, fig.cap="Diagnostic plots for the change-score model: residual Q-Q and residuals vs. fitted."}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
qqnorm(residuals(fit), main = "Normal Q-Q", pch = 19, col = navy, cex = .6)
qqline(residuals(fit), col = terra, lwd = 2)
plot(fitted(fit), residuals(fit), main = "Residuals vs Fitted",
     xlab = "Fitted", ylab = "Residual", pch = 19, col = navy, cex = .6)
abline(h = 0, lty = 2, col = gray)
```
